Towards Practical First-Order Model Counting

### 2024-04-19 doc/paper.tex
Section 5

We write $\mathbb{N}_{0}$ for the set of non-negative integers. In the context
of both algebra and logic, we write $C[x \mapsto y]$ for $C$ with all
occurrences of $x$ replaced with $y$.

@@

\begin{definition}
  Let $f$ be a function with signature $f(\mathbf{x})$. Then a \emph{base case}
  of $f$ is a function call $f(\mathbf{y})$ such that either $y_{i} = x_{i}$ or
  $y_{i} \in \mathbb{N}_{0}$ for all $i$, and the latter case applies at least
  once.
\end{definition}

@@

\section{Generating C++ Code (TODO)}\label{sec:cpp}

\todo[inline]{Remove mentions of LHS and RHS.}

The target is to generate C++ code that can evaluate numerical values of the
model counts based on the equations generated by \CompileWithBaseCases. The
translation of a set $\mathcal{E}$ of equations into a C++ program works as
follows.

First, we create a cache for each function in $\mathcal{E}$. This is implemented
as a multi-dimensional vector containing objects of \texttt{class cache\_elem}
defined as shown in the example code. The default initialization of this object
is to $-1$ which is useful for recognizing unevaluated cases.

Next, we create a function definition for the LHS of each equation in
$\mathcal{E}$, including all functions and base cases. The signatures of these
functions is decided as follows. A function call containing only variable
arguments is named as the function itself, and ones with constants in their
arguments are suffixed with a string that contains \texttt{'x'} at the $i$th
place if the $i$th argument is variable and the $i$th argument if that argument
is a constant. For example, $f(x_{1}, x_{2}, x_{3})$ is declared as \texttt{int
  f(int x1, int x2, int x3);} and $f(1, x_{2}, x_{3})$ is declared as
\texttt{int f\_1xx(int x2, int x3);} (the constant arguments are removed from
the signature).

The RHS of each equation in $\mathcal{E}$ is used to define the body of the
equation corresponding to the LHS of that equation. The function body (for a
function \texttt{func} corresponding to equation $e$) is formed as follows.

First, we check if the evaluation is already present in the cache. If so, then
we return the cache element. The cache accesses are done using the
\texttt{get\_elem} function (definition given in the example), which resizes the
cache if the accessed index is out of range.

Second, if the element is absent, then we decide if the arguments corresponding
to $e$ or one of the functions corresponding to the base cases, based on the
value of the arguments. If it corresponds to the base cases, then we directly
call the base case function and return its value. Else, we evaluate the value
using the RHS, store the evaluated value in the cache and return the evaluated
value. Note that in this step, we only call the base case function with one more
constant argument than \texttt{func}. For example, \texttt{f0(x, y)} would call
\texttt{f0\_0x(y)} if $x = 0$ and \texttt{f0\_x0(x)} if $y = 0$.

Third, to translate the RHS, we convert $\sum_{x=a}^{b} \expr{}$ to
\begin{lstlisting}[escapeinside={(*}{*)}]
([y,z,...](){
    int sum = 0;
    for(int x = a; x <= b; x++)
        sum += (*$\expr{}$*);
    return sum;
})()
\end{lstlisting}
where $y, z, \dots$ are the free variables present in $\expr{}$.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add some complete examples of C++ programs in the supplementary
          material.
          \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
            \item Something to think about that could inspire some kind of
                  theorems: how can I assume that a domain is empty if there are
                  constants associated with it?
          \end{itemize}
  \end{itemize}
}

>>>>

We write $\mathbb{N}_{0}$ for the set of non-negative integers. In the context
of both algebra and logic, we write $S\sigma$ for a \emph{substitution} $\sigma$
applied to an expression $S$, where
$\sigma = [x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$
denotes the replacement of all occurrences of $x_{i}$ with $y_{i}$ for all
$i = 1, \dots, n$.

@@

\begin{definition}\label{def:basecase}
  Let $f(\mathbf{x})$ be a function call where each $x_{i}$ is either a constant
  or a variable (note that signatures are included in this definition). Then a
  function call $f(\mathbf{y})$ is a \emph{base case} of $f(\mathbf{x})$ if
  $f(\mathbf{y}) = f(\mathbf{x})\sigma$, where $\sigma$ is a substitution that
  replaces one or more $x_{i}$ with a constant.
\end{definition}

@@

\todo[inline]{Somewhere: mention that we're primarily concerned with computing
  unweighted model counts, but everything trivially extends to the weighted
  setting as well.}

\section{Generating C++ Code}\label{sec:cpp}

In this section, we describe the last step of \Cranetwo{} outlined in
\cref{fig:overview}: translating the set of equations $\mathcal{E}$ produced by
\CompileWithBaseCases into C++ code. The produced C++ program can then be
compiled and executed with different command-line arguments to compute the model
count of the formula for different (combinations of) domain sizes.

Each equation in $\mathcal{E}$ is compiled into a C++ function, together with
its own cache for memoisation. Let
$e = (f(\mathbf{x}) = \expr{}) \in \mathcal{E}$ be an arbitrary equation, and
let $\mathbf{c} \in \mathbb{N}_{0}^{n}$ represent the arguments of the
corresponding C++ function. The implementation of $e$ consists of three parts.
First, we check whether $\mathbf{c}$ is already in the cache of $e$ (in which
case the cached value is returned). Second, for each base case $f(\mathbf{y})$
of $f(\mathbf{x})$ (as in \cref{def:basecase}), we check whether $\mathbf{c}$
\emph{matches} $\mathbf{y}$, i.e., $c_{i} = y_{i}$ whenever
$y_{i} \in \mathbb{N}_{0}$. In this case, $\mathbf{c}$ is redirected to the C++
function that corresponds to the definition of base case $f(\mathbf{y})$.
Finally, if the above cases fail, we evaluate $\mathbf{c}$ according to
$\expr{}$, store the result in the cache, and return it.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add some complete examples of C++ programs in the supplementary
          material.
          \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
            \item Something to think about that could inspire some kind of
                  theorems: how can I assume that a domain is empty if there are
                  constants associated with it?
            \item consider turning small sections (e.g., Section 5) into
                  subsections of even paragraphs.
          \end{itemize}
  \end{itemize}
}
